pirouette: the error BEAST2 makes in inferring a phylogeny

### 2019-01-21 article.tex
First useable PDF

@@

Phylogenies are commonly used to explore evolutionary hypotheses.
%Not only can phylogenies show us how species (or other
%evolutionary units) are related to each other, 
%but we can also estimate relevant parameters such as extinction and 
%speciation rates from them.
%There are many phylogenetics tools available to obtain an estimate 
%of the phylogeny of a given set of species. 
%BEAST2 (\cite{bouckaert2014beast}) is one of the most widely used ones.
%It uses a Bayesian statistical framework to estimate 
%the joint posterior distribution of phylogenies and model parameters, 
%from one or more DNA, RNA or amino acid alignments (see figure 1 
%for an overview of the workflow).

@@

\begin{figure}[h]
  \includegraphics[width=\textwidth]{overview.png}
\end{figure}

@@

\verb;piouette; can be installed easily from CRAN:

@@

This code will create a (temporary) BEAST2 configuration file,
from the FASTA file with name \verb;anthus_aco.fas; (which
is supplied with the package, from \cite{VanEls2018}), 
using the same default settings as BEAUti, which are, 
among others, a Jukes-Cantor site model, a strict clock, and a Yule birth tree prior.
\verb;pirouette; will then execute BEAST2 using that file, and
parses the output. The returned data structure, named \verb;out;, 
is a list of parameter estimates (called \verb;estimates;), posterior 
phylogenies (called \verb;anthus_aco_trees;, named after
the alignment's name) and MCMC operator performance (\verb;operators;).
An example of using a different site model, clock model 
and tree prior is:

\begin{lstlisting}[language=R, floatplacement=H]
out <- bbt_run(
  fasta_filenames = "anthus_aco.fas",
  site_models = create_hky_site_model(),
  clock_models = create_rln_clock_model(),
  tree_priors = create_bd_tree_prior()
)
\end{lstlisting}
This code uses an HKY site model, a relaxed log-normal clock model and a 
birth-death tree prior, each with their default settings in BEAUti.
Table \ref{tab:functions} shows an overview of all functions to 
create site models, clock models and tree priors.
Note that the arguments' names \verb;site_models;, \verb;clock_models; 
and \verb;tree_priors; are plural, as each of these
can be (a list of) one or more elements. Each of these arguments must 
have the same number of elements, so that each alignment has its
own site model, clock model and tree prior. 
An example of two alignments, each with its own site model, is:

\begin{lstlisting}[language=R, floatplacement=H]
out <- bbt_run(
  fasta_filenames = c(
    "anthus_aco.fas", 
    "anthus_nd2.fas"
  ),
  site_models = list(
    create_tn93_site_model(), 
    create_gtr_site_model()
  )
)
\end{lstlisting}
\verb;pirouette; also uses the same default prior distributions as BEAUti 
for each of the site models, clock models and tree priors. 
For example, by default, a Yule tree prior assumes that the birth rate 
follows a uniform distribution, 
from minus infinity to plus infinity. 
One may prefer a different ddistribution instead. 
Here is an example how to specify an exponential distribution for
the birth rate in a Yule tree prior in \verb;pirouette;:

\begin{lstlisting}[language=R, floatplacement=H]
out <- bbt_run(
  fasta_filenames = "anthus_aco.fas",
  tree_priors = create_yule_tree_prior(
    birth_rate_distr = create_exp_distr()    
  )
)
\end{lstlisting}
In this same example, one may specify
the initial shape parameters of the exponential distribution.
In BEAST2's implementation, an exponential distribution 
has one shape parameter: its mean, which can be set to any
value with \verb;BEAUti;. To set the 
mean value of the exponential distribution to a 
fixed (non-estimated) value, do:

\begin{lstlisting}[language=R, floatplacement=H]
out <- bbt_run(
  fasta_filenames = "anthus_aco.fas",
  tree_priors = create_yule_tree_prior(
    birth_rate_distr = create_exp_distr(
      mean = create_mean_param(
        value = 1.0, 
        estimate = FALSE
      )
    )    
  )
)
\end{lstlisting}
\verb;pirouette; also supports node dating. Like BEAUti, one
can specify Most Recent Common Ancestor ('MRCA') priors.
An MRCA prior allows to specify taxa having a common ancestor,
including a distribution for the date of that ancestor.
With \verb;pirouette;, this is achieved as follows:

\begin{lstlisting}[language=R, floatplacement=H]
out <- bbt_run(
  fasta_filenames = "anthus_aco.fas",
  mrca_priors = create_mrca_prior(
    taxa_names = sample(get_taxa_names("anthus_aco.fas"), size = 2),
    alignment_id = get_alignment_id("anthus_aco.fas"),
    is_monophyletic = TRUE,
    mrca_distr = create_normal_distr(
      mean = create_mean_param(value = 15.0, estimate = FALSE),
      sigma = create_sigma_param(value = 0.025, estimate = FALSE)
    )
  )
)
\end{lstlisting}
Instead of dating the ancestor of two random taxa, any subset of taxa can be selected,
and multiple sets are allowed.
\verb;pirouette; allows for the same core functionality as Tracer to show the values of the parameter estimates sampled
in the BEAST2 run. This is called the "trace" (hence the name).
The start of the trace, called the "burn-in", is usually discarded, as an MCMC 
algorithm (such as used by BEAST2) first has to converge to
its equilibrium and hence the parameter estimates are not 
representative. By default, Tracer discards the first 10\% of all 
the parameter estimates. 
To remove a 20\% burn-in from all parameter estimates 
in \verb;pirouette;, the following code can be used:

\begin{lstlisting}[language=R, floatplacement=H]
traces <- remove_burn_ins(
  traces = out$estimates, 
  burn_in_fraction = 0.2
)
\end{lstlisting}
Tracer shows the ESSes of each posterior's variables.
These ESSes are important to determine the strength of the
inference. As a rule of thumb, an ESS of 200 is acceptable 
for any parameter estimate.
To calculate the effective sample sizes (of all estimated variables) in \verb;pirouette;:

\begin{lstlisting}[language=R, floatplacement=H]
esses <- calc_esses(
  traces = traces, 
  sample_interval = 1000
)
\end{lstlisting}
Tracer displays multiple summary statistics for each
estimated variable: the mean and its standard error, standard deviation,
variance, median, mode, geometric mean, 95\% highest posterior density interval, 
auto-correlation time and effective sample size. It displays these statistics per
variable. In \verb;pirouette;, these summary statistics are collected for
all estimated parameters at once:

\begin{lstlisting}[language=R, floatplacement=H]
sum_stats <- calc_summary_stats(
  traces = traces, 
  sample_interval = 1000
)
\end{lstlisting}
\verb;pirouette; allows for the same functionality as DensiTree.
DensiTree displays the phylogenies in a posterior at the same
time scale, drawn one over one another, allowing to see the uncertainty in
topology and branch lengths. 
The posterior phylogenies are stored as \verb;anthus_aco_trees; in the object \verb;out;,
and can be plotted as follows:

\begin{lstlisting}[language=R, floatplacement=H]
plot_densitree(phylos = out$anthus_aco_trees)
\end{lstlisting}
Instead of running the full pipeline, 
\verb;pirouette; also allows to only create a BEAST2 configuration file.
To create a BEAST2 configuration file, with all settings to default, use:

\begin{lstlisting}[language=R, floatplacement=H]
create_beast2_input_file(
  input_filenames = pirouette::get_pirouette_path("anthus_aco.fas"),
  output_filename = "beast2.xml"
)
\end{lstlisting}
This file can then be loaded and edited by BEAUti, 
run by BEAST2, or run by \verb;pirouette;:

\begin{lstlisting}[language=R, floatplacement=H]
run_beast2(
  input_filename = "beast2.xml",
  output_log_filename = "run.log",
  output_trees_filenames = "posterior.trees",
  output_state_filename = "final.xml.state"
)
\end{lstlisting}
\verb;run_beast2; is a function that only runs BEAST2, 
and does not parse the output files (unlike \verb;bbt_run;). 
In the example above, 
we specify the names of the desired BEAST2 
output files explicitly, and these will be created in the
R working directory, after which they can be inspected 
with other tools, or used to continue a BEAST2 run.
When the names of these files are not specified, 
both \verb;bbt_run; and \verb;run_beast2;
put these files in the default temporary folder (as
obtained from \verb;temp.dir();) to keep
the working directory clean of intermediate files.

@@

\verb;pirouette; is free, libre and open source software available at 
\url{http://github.com/richelbilderbeek/pirouette}
and is licensed under the GNU General Public License v3.0.
\verb;pirouette; uses the Travis CI (\url{https://travis-ci.org})
continuous integration service, which is known to significantly 
increase the number of bugs exposed (\cite{vasilescu2015}) and increases
the speed at which new features are added (\cite{vasilescu2015}).
\verb;pirouette; has a 100\% code coverage, which correlates with 
code quality (\cite{horgan1994,del1995correlation}). 
\verb;pirouette; follows Hadley Wickham's style guide (\cite{style_guide}), 
which improves software quality (\cite{fang2001}).
\verb;pirouette; depends on multiple packages, which are 
\verb;ape; (\cite{APE}), 
\verb;babette; (\cite{babette}),
\verb;ggplot2; (\cite{ggplot2}),
\verb;knitr; (\cite{knitr}),
\verb;mcbette; (\cite{mcbette}),
\verb;phangorn; (\cite{phangorn}),
\verb;rmarkdown; (\cite{rmarkdown}),
\verb;stringr; (\cite{stringr}),
\verb;testit; (\cite{testit}) and 
\verb;usethis; (\cite{usethis}).

>>>>

% sidewaysfigure
\usepackage{rotating}

@@

Phylogenies are commonly used to explore evolutionary hypotheses.

%Not only can phylogenies show us how species (or other
%evolutionary units) are related to each other, 
%but we can also estimate relevant parameters such as extinction and 
%speciation rates from them.
%There are many phylogenetics tools available to obtain an estimate 
%of the phylogeny of a given set of species. 
%BEAST2 (\cite{bouckaert2014beast}) is one of the most widely used ones.
%It uses a Bayesian statistical framework to estimate 
%the joint posterior distribution of phylogenies and model parameters, 
%from one or more DNA, RNA or amino acid alignments (see figure 1 
%for an overview of the workflow).

@@

\begin{sidewaysfigure}
  \centering
  \includegraphics[width=\textwidth]{overview.png}
  \caption{pirouette pipeline}
  \label{fig:pipeline}
\end{sidewaysfigure}

@@

\verb;pirouette; can be installed easily from CRAN:

@@

@@

\verb;pirouette; is free, libre and open source software available at 
\url{http://github.com/richelbilderbeek/pirouette}
and is licensed under the GNU General Public License v3.0.
\verb;pirouette; uses the Travis CI (\url{https://travis-ci.org})
continuous integration service, which is known to significantly 
increase the number of bugs exposed (\cite{vasilescu2015}) and increases
the speed at which new features are added (\cite{vasilescu2015}).
\verb;pirouette; has a 100\% code coverage, which correlates with 
code quality (\cite{horgan1994,del1995correlation}). 
\verb;pirouette; follows Hadley Wickham's style guide (\cite{style_guide}), 
which improves software quality (\cite{fang2001}).
\verb;pirouette; depends on multiple packages, which are 
\verb;ape; (\cite{APE}), 
\verb;babette; (\cite{bilderbeek2018babette}),
\verb;ggplot2; (\cite{ggplot2}),
\verb;knitr; (\cite{knitr}),
\verb;mcbette; (\cite{mcbette}),
\verb;phangorn; (\cite{phangorn}),
\verb;rmarkdown; (\cite{rmarkdown}),
\verb;stringr; (\cite{stringr}),
\verb;testit; (\cite{testit}) and 
\verb;usethis; (\cite{usethis}).
